% !TEX root =  ../thesis.tex

This work couldn't be completed without the work, help, advice, presence of persons other than myself. I will try to appropriately thank all of them.

I want to express my sincerest gratitude to my advisor Lenore Zuck. Thank you also for giving to me the opportunity to work and contribute to this project. I would like to thank the people that I had the honor to work with in the past year: Niko Zarzani, Professor Rigel Gjomemo, Phu Phung, and Professor and advisor Venkat Venkatakrishnan. Heartfelt thanks to Kedar Namjoshi, without whose constant support, advice and guidance this work wouldn't have seen the light.

I would like to thank Professors Stefano Zanero, Marco Santambrogio, Pier Luca Lanzi at Politecnico di Milano, for helping and inspiring before and during the master's degree. Same Sincere thanks to Lynn Thomas, for assisting me whenever the need.

I want to thank my housemates Claudio, Daniele, Giovanni, Gabriele, Luca for tolerating me during the hardest moments. Thank you, Dasha, the main source of happiness and mental stability in my life. A lot of thanks to my family, who gave me the opportunity to be here. Thanks Papà, Mamma e Claudia, I missed you all the time.

Thanks to Michele Spagnuolo for sharing his Thesis template with me to use it in this very work.

Last but not least, I owe a lot to the Computer Science community in general, and to some amazing personalities in particular. Even if you don't know me, Thank you Linus Torvalds, Chris Lattner, Jimbo Wales, Jeff Atwood, Joel Spolsky, Leonardo de Moura, LLVM and CLANG community members. I will repay my debt somehow.

For all the other friends, professors and colleagues that I didn't name in this very brief list: thank you all.
\\\\

\begin{flushright}
{\large GT}
\end{flushright}
